% !BIB TS-program = biber

\RequirePackage[l2tabu,orthodox]{nag}

% TODO: decide if one-sided/two-sided
%\documentclass[headsepline,footsepline,footinclude=false,fontsize=11pt,paper=a4,listof=totoc,bibliography=totoc,BCOR=12mm,DIV=12]{scrbook} % two-sided
\documentclass[headsepline,footsepline,footinclude=false,oneside,fontsize=11pt,paper=a4,listof=totoc,bibliography=totoc]{scrbook} % one-sided

% TODO: change thesis information
\newcommand*{\getUniversity}{Technische Universität München}
\newcommand*{\getFaculty}{Informatics}
\newcommand*{\getDegree}{Informatics}
\newcommand*{\getSchool}{Computation, Information and Technology}
\newcommand*{\getTitle}{Performance Prediction of Iterator-Based Database Systems with Decision Trees}
\newcommand*{\getTitleGer}{Performancevorhersage iteratorbasierter Datenbanksysteme mit Entscheidungsbäumen}
\newcommand*{\getAuthor}{Hoang-Nam Tran}
\newcommand*{\getDoctype}{Bachelor's Thesis}
\newcommand*{\getSupervisor}{Prof. Dr. Thomas Neumann}
\newcommand*{\getAdvisor}{Maximilian Rieger, M.Sc.}
\newcommand*{\getKeywords}{keyword;another keyword;one more}
\newcommand*{\getSubmissionDate}{03.05.2026}
\newcommand*{\getSubmissionLocation}{Munich}

% TODO: change citation style in settings
\input{settings}

\begin{document}

% Set page numbering to avoid "destination with the same identifier has been already used" warning for cover page.
% (see https://en.wikibooks.org/wiki/LaTeX/Hyperlinks#Problems_with_Links_and_Pages).
\pagenumbering{alph}
\input{pages/cover}

\frontmatter{}

\input{pages/title}
\input{pages/disclaimer}
\input{pages/acknowledgments}
\input{pages/abstract}
\microtypesetup{protrusion=false}
\tableofcontents{}
\microtypesetup{protrusion=true}

\mainmatter{}

\input{chapters/01_introduction}
% TODO: add more chapters here
\input{chapters/02_theoretical_background}
\input{chapters/03_methodology}
\input{chapters/04_implementation}
\input{chapters/05_evaluation}
\input{chapters/06_future_work}
\input{chapters/07_conclusion}

\appendix{}



\microtypesetup{protrusion=false}



\listoffigures{}
\listoftables{}
\microtypesetup{protrusion=true}

\addchap{Declaration on the Use of AI-Assisted Tools}

We used AI tools in two capacities during this thesis.

For implementation, we used Claude to generate code.
Generated code was reviewed, tested, and adapted to the specific requirements of this project.

Regarding writing, we used Claude as an assistant for drafting, structuring, and rephrasing text.
All written content was critically reviewed, verified, and revised for accuracy and coherence.

The research design, methodology, experiments, analysis, and conclusions are our own.

\printbibliography{}

\end{document}
