% !TeX root = ../main.tex
% Add the above to each chapter to make compiling the PDF easier in some editors.

\chapter{Methodology}\label{chapter:methodology}

\section{Problem Statement and Prediction Target}

In this thesis, we address the transfer of the T3 runtime prediction approach from Umbra to PostgreSQL parsed plans.  
The central methodological goal is to preserve T3’s core modeling principles.
Upon that, we adapt plan conversion and feature extraction to PostgreSQL-specific plan information.

The final prediction target is total query runtime.  
Each query plan is decomposed into pipelines and the model predicts a duration for each pipeline.
The predicted query runtime is the sum of these per-pipeline predictions.

\section{Data and Preprocessing}
Regarding the training data, we use the public Zero-Shot benchmark data introduced with the Zero-Shot Cost Model.  
The benchmark is particularly suitable for this thesis because it consists of real-world datasets (e.g., IMDB) and synthetically generated workloads designed to cover diverse query structures.
This includes \emph{SPAJ} (Select-Project-Aggregate-Join) queries with conjunctive predicates.
Overall, the benchmark contains 20 databases with large variation in tables, columns, and foreign-key relationships \parencite{zeroshot}.

Having this multitude of databases allows us to test the generalization capabilities of our model.
Similar to the T3 paper, we can train the model on all but one databases and test on the remaining one.
Moreover, as the widely used JOB workload is included, we can compare our performance predictions to other implementations later on.
Further, PostgreSQL is the intended database system of our implementation, which avoids the effort of building a separate PostgreSQL benchmark.

Following the benchmark construction described in the paper, queries are generated per database and executed on a PostgreSQL DBMS (v12) on c8220 nodes on the CloudLab platform.
Then, they are stored as raw traces (i. e., the output of \texttt{EXPLAIN ANALYZE}).
Next, the authors preprocess the queries to remove very long-running queries (queries above 30\,s are excluded) to keep the benchmark practical and reproducible \parencite{zeroshot}.
A further preprocessing step is to parse the raw traces into structured plan representations.
These \texttt{parsed\_plans} (15000 per database) contain the operator nodes found in the raw traces, but organize them in a tree structure with node-wise attributes.
For every node, e.g., operator type, estimated/actual cardinalities, widths, and runtimes are stored in the \texttt{plan\_parameters} field.

Figure~\ref{fig:parsed-plan-hash-join-cutout} shows such a structured plan representation for a \texttt{Hash Join} and its children.
Here, also filter predicates can be retrieved from the \texttt{filter\_columns} field.
Furthermore, PostgreSQL-specific fields like \texttt{workers\_planned} (planned parallel workers for the node) are available.


The tree structure can be used to create pipelines and the node information will be used to train the model.
In the upcoming sections, we will describe these key implementation details.

Consequently, the parsed plans offer a good starting point for our pipeline and we can begin without further preprocessing.

\begin{figure}[htpb]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
    planop/.style={draw, rounded corners, fill=gray!10, align=left, inner sep=6pt, text width=0.46\linewidth, font=\small}
  ]
    \node[planop] (join) at (0,0) {%
      \texttt{op\_name: Hash Join}\\
      \texttt{est\_startup\_cost: 120636.97}\\
      \texttt{est\_cost: 319835.72}\\
      \texttt{est\_card: 14988952.0}\\
      \texttt{est\_width: 0.0}\\
      \texttt{act\_startup\_cost: 609.469}\\
      \texttt{act\_time: 1473.528}\\
      \texttt{act\_card: 6611280.0}\\
      \texttt{filter\_columns: \{column: 56, operator: =,}\\
      \texttt{\ \ literal: corporation.region\_no,\ literal\_feature: 0\}}\\
      \texttt{act\_children\_card: 254280000.0}\\
      \texttt{est\_children\_card: 1176670000.0}\\
      \texttt{workers\_planned: 2}
    };
    \node[planop] (left) at (-6.3,-8.2) {%
      \texttt{op\_name: Merge Join}\\
      \texttt{est\_startup\_cost: 120548.72}\\
      \texttt{est\_cost: 126900.67}\\
      \texttt{est\_card: 470668.0}\\
      \texttt{est\_width: 12.0}\\
      \texttt{act\_startup\_cost: 608.149}\\
      \texttt{act\_time: 638.269}\\
      \texttt{act\_card: 101712.0}\\
      \texttt{act\_children\_card: 795865104.0}\\
      \texttt{est\_children\_card: 18939960.0}\\
      \texttt{workers\_planned: 2}
    };
    \node[planop] (hash) at (6.3,-8.2) {%
      \texttt{op\_name: Hash}\\
      \texttt{est\_startup\_cost: 57.0}\\
      \texttt{est\_cost: 57.0}\\
      \texttt{est\_card: 2500.0}\\
      \texttt{est\_width: 4.0}\\
      \texttt{act\_startup\_cost: 1.208}\\
      \texttt{act\_time: 1.209}\\
      \texttt{act\_card: 2500.0}\\
      \texttt{act\_children\_card: 2500.0}\\
      \texttt{est\_children\_card: 2500.0}\\
      \texttt{workers\_planned: 2}
    };
    \node[planop] (scan) at (6.3,-15.0) {%
      \texttt{table: 2}\\
      \texttt{op\_name: Seq Scan}\\
      \texttt{est\_startup\_cost: 0.0}\\
      \texttt{est\_cost: 57.0}\\
      \texttt{est\_card: 2500.0}\\
      \texttt{est\_width: 4.0}\\
      \texttt{act\_startup\_cost: 0.026}\\
      \texttt{act\_time: 0.765}\\
      \texttt{act\_card: 2500.0}\\
      \texttt{filter\_columns: \{\}}\\
      \texttt{act\_children\_card: 1}\\
      \texttt{est\_children\_card: 1}\\
      \texttt{workers\_planned: 2}
    };

    \draw[-latex, thick] (left) -- (join);
    \draw[-latex, thick] (hash) -- (join);
    \draw[-latex, thick] (scan) -- (hash);
  \end{tikzpicture}
  }
  \caption[Top-node cutout with Hash Join]{Top-node tree cutout of one \texttt{parsed\_plan} with a \texttt{Hash Join} from \path{credit/workload_100k_s1_c8220}. Child relationships are shown by arrows,
   and each node box contains selected \texttt{plan\_parameters} of that node for the displayed cutout.}
  \label{fig:parsed-plan-hash-join-cutout}
\end{figure}



\section{Implementation}\label{sec:methodology-implementation}


This section describes how a single parsed plan from the Zero-Shot benchmark is turned into a T3-style representation including pipelines.
Then, we derive fixed-length feature vectors for each pipeline.

The procedure follows the pipeline prediction idea of T3~\parencite{t3}, but also replaces Umbra-specific features with properties that PostgreSQL exposes in the \texttt{parsed\_plans}.
Once all the plans are converted and the feature vectors are created, we can train the model.

\subsection{Mapping Parsed PostgreSQL Plans to the T3 Plan Format}

We start from one element, i. e., one plan, of the \texttt{parsed\_plans} array in the benchmark JSON files.
An object in the array is a rooted tree: every node carries \texttt{plan\_parameters} (operator name, estimated and actual cardinalities, 
tuple width, planner costs, optional filter predicates, timings, and further fields) and a list \texttt{children} of child subtrees.


In \texttt{zeroshot\_to\_t3.py}, we convert this tree recursively into the same nested operator structure
 that the original T3 implementation expects
 for Umbra plans~\parencite{t3}.
PostgreSQL operator names are mapped to abstract operators (\texttt{tablescan}, \texttt{join}, \texttt{sort}, and \texttt{groupby} nodes).
The PostgreSQL-specific temporary operators  \texttt{hash} and \texttt{materialize} are mapped to the abstract operator \texttt{temp}.

Regarding joins, the child order is normalized so that it matches T3's pipeline rules: i. e., the PG-build subtree becomes the \emph{left} child and the PG-probe subtree the \emph{right} child.

Since the feature vector uses the PostgreSQL-native features and operators of the \texttt{parsed\_plans}, we need to keep the original, not converted \texttt{plan\_parameters} available for the feature extraction.
Hence, the converted abstract operator names are only used to match the internal T3 representation for pipeline assignment, but not for the feature extraction.

Consequently, every converted node additionally stores a copy of the original \texttt{plan\_parameters} under a dedicated \texttt{pg} payload.
Training and inference, that use PostgreSQL-native features, read from this payload and from the pipeline layout, which will be described in the next sections.

T3 also requires database schema information. However, the Zero-Shot data does not provide this information, so a minimal placeholder schema is used instead.

\subsection{Pipeline Decomposition}

Umbra plans group operators into \emph{pipelines}: sets of operators that can produce tuples in a pipelined fashion, separated by \emph{pipeline breakers}. 
Namely these are operators/stages, where data is materialized. For instance, sorting, grouping, or materialization, but also the build side of a hash join~\parencite{t3}.

Since PostgreSQL plans do not provide such a pipeline breakdown, we need to derive it from the converted tree.
We assign each node to a pipeline by walking the converted tree with the same high-level rules as in the T3 work.
Thus, PG's sort and aggregate nodes break pipelines; a PG-materialize node is treated as a breaker.
Joins place the build side in one pipeline and keep the join together with the probe side in the other pipeline.
The pipeline assignment is stored in the \texttt{analyzePlanPipelines} list of the converted plan.

For every pipeline, we derive start and stop times.
They stem from the per-node actual timings in \texttt{plan\_parameters} and are accumulated, so that pipeline durations are available for further usage during training.

\subsection{PostgreSQL-Native Feature Vectors}

The original T3 feature mapper for Umbra builds a feature vector from execution phases within each pipeline (e.g.\ scan, build, probe phases) and from cardinalities and tuple sizes attached to those phases~\parencite{t3}.
As PostgreSQL plans contain different data, we use \texttt{pg\_features.py} to construct the feature vector for each pipeline after the aforementioned pipeline decomposition.
It becomes clear, why the \texttt{plan\_parameters} are kept in the converted tree (\texttt{pg} payload entry), as we use them for the feature extraction.

Once the feature vector is constructed, we can train the model.
Our feature vector differs from the original T3 feature vector in the following ways:



\begin{enumerate}
  \item \textbf{PostgreSQL-native operator features.} The \texttt{parsed\_plans} expose PostgreSQL-native operator nodes. 
  Thus, we add entries for these PostgreSQL-specific operators in the feature vector.
  As no execution stages are available, we aggregate features only per operator.
  Hence, we do not include the stage-based features of the original T3 feature mapper. This avoids inventing synthetic stages for every node.
  \item \textbf{PostgreSQL planner cost estimates.} We include sums and extrema of PostgreSQL's \texttt{est\_cost} and \texttt{est\_startup\_cost} from \texttt{plan\_parameters} in the feature vector.
  These are optimizer-internal estimates of work and startup cost and are an additional PostgreSQL-specific signal that is not available in the Umbra feature vector.
  Consequently, we can include this information in the feature vector.
  \item \textbf{Planned parallelism.} Another PostgreSQL-specific signal is the planned parallelism. So, we take the sum of \texttt{workers\_planned} over all operators in the pipeline.
  Hereby, we include a parallelism signal in our feature vector.
  \item \textbf{Sizes.} T3 uses tuple sizes in its feature vector. This information is provided by the Umbra plans.
  However, the parsed PostgreSQL do not contain a perfect equivalent. 
  Instead, they report a single width estimate per node (\texttt{est\_width}).
  Thus, an average of \texttt{est\_width} over all operators in the pipeline is used to represent sizes in the feature vector, replacing the tuple size features of T3.
\end{enumerate}

In the Umbra feature vector of T3, rows are indexed by both operator type and execution phase (scan, build, probe, \ldots) like \texttt{HashJoin\_Probe\ldots}~\parencite{t3}.
Now with the PostgreSQL implementation, we miss out on the execution phase dimension.

Whereas not an equivalent substitute, we replace this information with a pipeline-level dimension and keep the operator-level features for the individual operators.
The pipeline-level block produces additional metadata in a compact manner, e. g., by using averages and sums over operators.
For instance, we consider planned parallelism, width, and cost estimates per pipeline and not per operator.
Subsequently, the feature vector stays more compact.

Each pipeline's feature vector consists of two concatenated blocks, as Figure~\ref{fig:pg-feature-vector-snippet} shows:

\begin{enumerate}
  \item \textbf{Pipeline-level aggregates.}
  From the available \texttt{pg} payload, we try to extract as distinct information as possible.

  Therefore, we start by summarizing different information categories per pipeline with sum, max, and average statistics.
  First, the \texttt{pg\_card\_...} fields capture output cardinality summaries.
  
  Insights about PostgreSQL-specific cost, parallelism, and width features are captured in the
  \texttt{pg\_est\_cost\_...},
  \texttt{pg\_est\_startup\_...},
  \texttt{pg\_workers\_planned\_sum} 
  and \texttt{pg\_est\_width\_avg} fields.
  
  Further, the \texttt{pg\_num\_(scan/join/sort/etc.)} family counts operator classes.
  
  Moreover, the \texttt{pg\_scan\_has\_filter}
  and \texttt{pg\_filter\_(and/or/compare/like/in)} counters enumerate scan predicate attributes.
  
  Finally, pipeline-context features describe structure using
  \texttt{pg\_pipeline\_num\_ops} for instance.
  \item \textbf{Operator-level block.} 
  For every operator that occurs in \texttt{parsed\_plans}, we assign a slice in the feature vector.
  If multiple operators of the same type occur in the pipeline, we sum the features over all occurences.
  In figure~\ref{fig:pg-feature-vector-snippet}, the excerpt of the operator-type block for the \texttt{Nested\_Loop join} operator is depicted.
  
  Depending on the operator type, we include different features. This feature set is derived from T3's Umbra implementation ~\parencite{t3}.
  For example, a join would include different features than a scan, like in the Umbra/T3 approach.
  
  Regarding the calculations of the features, we follow the Umbra/T3 implementation where adequate.
  A deviation would be, e. g, how expression percentages are formed due to the different nature of PostgreSQL's filter predicates (CITE).
  
  Finally, we include the operator count "\texttt{const}", which exists for every operator type in both our and the Umbra/T3 implementation ~\parencite{t3}.
\end{enumerate}

In total, the feature vector is fixed and has 153 dimensions (23 pipeline-level features and 130 operator-type features).
Empty features are zeroed to keep the feature vector dimension.

Regarding the cardinalities, either the actual or the estimated cardinalities are used in the feature vector, depending on which is selected in training.
However, even if estimates are used, the row names are still called "actual" cardinalities.
This caveat is a design decision to keep the row names completely uniform.

Once the feature vectors for all pipelines are constructed, we get the feature matrix for the whole plan.
We proceed to train the model in the next section.


\begin{figure}[htpb]
  \centering
  \footnotesize
  \setlength{\tabcolsep}{5pt}
  \begin{tabular}{@{}l r@{}}
    \toprule
    \multicolumn{2}{@{}l}{\emph{Pipeline-level block (excerpt; first 12 of 23 features)}} \\
    \midrule
    \texttt{pg\_card\_sum} & 6541.0 \\
    \texttt{pg\_card\_max} & 6537.0 \\
    \texttt{pg\_est\_cost\_sum} & 16882.95 \\
    \texttt{pg\_est\_cost\_max} & 16877.3 \\
    \texttt{pg\_est\_startup\_sum} & 125.82 \\
    \texttt{pg\_workers\_planned\_sum} & 4.0 \\
    \texttt{pg\_est\_width\_avg} & 10.0 \\
    \texttt{pg\_num\_scan} & 1.0 \\
    \texttt{pg\_num\_join} & 1.0 \\
    \texttt{pg\_scan\_has\_filter} & 0.0 \\
    \texttt{pg\_filter\_compare\_count} & 0.0 \\
    \texttt{pg\_pipeline\_num\_ops} & 2.0 \\
    \midrule
    \multicolumn{2}{@{}l}{\emph{Operator-type block (excerpt; Nested\_Loop slice; 130 entries in full)}} \\
    \midrule
    \texttt{Nested\_Loop\_const} & 1.0 \\
    \texttt{Nested\_Loop\_in\_percentage} & 0.11473152822395594 \\
    \texttt{Nested\_Loop\_right\_card} & 98378.0 \\
    \texttt{Nested\_Loop\_out\_percentage} & 0.008260670032124828 \\
    \multicolumn{2}{@{}c@{}}{\scriptsize\textit{(remaining operator-type entries omitted)}} \\
    \bottomrule
  \end{tabular}
  \caption[Excerpt of a PostgreSQL-native pipeline feature vector]
  {Excerpts of the pipeline-level block (first twelve features) and of the \texttt{Nested\_Loop} operator entries in the operator-type block. 
  They are from one pipeline of query \texttt{index\_workload\_100k\_s2\_c8220\_890} on the TPC-H database. 
  The full vector has 153 dimensions; remaining pipeline and operator entries are omitted.}
  \label{fig:pg-feature-vector-snippet}
\end{figure}


\section{Training}

Concerning the training, we keep T3's LightGBM decision tree setup~\parencite{t3}.
Then, we use the newly constructed PostgreSQL-native feature matrix.
A main training difference is that we train directly on pipeline durations as labels rather than using tuple-based targets.

\subsection{Training Targets and Model Configuration}

Concerning the training labels, we use the observed pipeline duration.
In T3~\parencite{t3}, the training target is the per-tuple-time instead, which is the pipeline duration divided by a pipeline scan cardinality.
We will compare the per-tuple-time approach with the pipeline duration approach in the REFERENCEABLATIONSTUDY.

Regarding inference, our predicted query runtime is the sum of the predicted pipeline durations.


Our model uses gradient boosted decision trees as implemented in LightGBM~\parencite{lightgbm}. 
The booster is trained with a mean absolute percentage error (MAPE) objective on the training rows.
This is in line with the original T3 work~\parencite{t3}.

Furthermore, we use the same training parameters as in the original T3 work~\parencite{t3}.
Hence, the pipeline duration targets $t$ are transformed as $-\log(t)$ to stabilize training.
Also following T3, we cap the number of boosting rounds at 200 trees to mitigate overfitting.

\subsection{Data Splits and Reproducibility}

\paragraph{Generalization.}
We match the training intention of the Zero-Shot benchmark~\parencite{zeroshot} and of T3's cross-instance evaluation~\parencite{t3}.
Both aim to measure whether the trained model transfers to an unseen database instance.
Thus, we use a leave-one-out split on the \texttt{parsed\_plan} files:
We specify one instance as a holdout and use its \texttt{parsed\_plan} files as the test set.
All other instances form the training data.

Concretely, \texttt{training\_zeroshot\_tpch\_holdout.py} uses the \texttt{-\--holdout} argument to specify the holdout instance (default: \texttt{tpc\_h}).
Then, we perform a training for all test instances with \texttt{run\_all\_holdouts.py}.

\paragraph{Train and Validation Splits.}
Mirroring T3's train/validation split, we sample 80\% of the training data for training and 20\% for validation~\parencite{t3}.
The default seed used for shuffling the training data is 42.

\paragraph{Environment.}
Model training and evaluation for the results reported in this thesis were run on a personal laptop: an Apple MacBook Pro (16-inch, 2021) with the Apple M1 Pro processor (10 CPU cores, 16 GPU cores), 16\,GB memory, and a 512\,GB SSD, under macOS 15.7.3.

\section{Other Performance Prediction Implementations}

Next to already described T3 and Zeroshot...

\paragraph{T3-based Models with Different Feature Vectors.}

Wehrstein et al.~\parencite{jobcomplex} also adapted the Umbra T3 model for PostgreSQL.
They use a hybrid feature vector that takes the original Umbra feature vector, and maps matching PostgreSQL operators to Umbra operators.
On top of that, they add additional features that are specific to PostgreSQL, like the Merge Join operator.
However, they do not use fields that account for PostgreSQL-specific characteristics, like the planned workers or the estimated startup cost fields we use for our model ~\parencite{jobcomplex}.

\section{Section}
Citation test~\parencite{latex}.

Acronyms must be added in \texttt{main.tex} and are referenced using macros. The first occurrence is automatically replaced with the long version of the acronym, while all subsequent usages use the abbreviation.

E.g. \texttt{\textbackslash ac\{TUM\}, \textbackslash ac\{TUM\}} $\Rightarrow$ \ac{TUM}, \ac{TUM}

For more details, see the documentation of the \texttt{acronym} package\footnote{\url{https://ctan.org/pkg/acronym}}.
\subsection{Subsection}

See~\autoref{tab:sample}, \autoref{fig:sample-drawing}, \autoref{fig:sample-plot}, \autoref{fig:sample-listing}.

\begin{table}[htpb]
  \caption[Example table]{An example for a simple table.}\label{tab:sample}
  \centering
  \begin{tabular}{l l l l}
    \toprule
      A & B & C & D \\
    \midrule
      1 & 2 & 1 & 2 \\
      2 & 3 & 2 & 3 \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{figure}[htpb]
  \centering
  % This should probably go into a file in figures/
  \begin{tikzpicture}[node distance=3cm]
    \node (R0) {$R_1$};
    \node (R1) [right of=R0] {$R_2$};
    \node (R2) [below of=R1] {$R_4$};
    \node (R3) [below of=R0] {$R_3$};
    \node (R4) [right of=R1] {$R_5$};

    \path[every node]
      (R0) edge (R1)
      (R0) edge (R3)
      (R3) edge (R2)
      (R2) edge (R1)
      (R1) edge (R4);
  \end{tikzpicture}
  \caption[Example drawing]{An example for a simple drawing.}\label{fig:sample-drawing}
\end{figure}

\begin{figure}[htpb]
  \centering

  \pgfplotstableset{col sep=&, row sep=\\}
  % This should probably go into a file in data/
  \pgfplotstableread{
    a & b    \\
    1 & 1000 \\
    2 & 1500 \\
    3 & 1600 \\
  }\exampleA
  \pgfplotstableread{
    a & b    \\
    1 & 1200 \\
    2 & 800 \\
    3 & 1400 \\
  }\exampleB
  % This should probably go into a file in figures/
  \begin{tikzpicture}
    \begin{axis}[
        ymin=0,
        legend style={legend pos=south east},
        grid,
        thick,
        ylabel=Y,
        xlabel=X
      ]
      \addplot table[x=a, y=b]{\exampleA};
      \addlegendentry{Example A}
      \addplot table[x=a, y=b]{\exampleB};
      \addlegendentry{Example B}
    \end{axis}
  \end{tikzpicture}
  \caption[Example plot]{An example for a simple plot.}\label{fig:sample-plot}
\end{figure}

\begin{figure}[htpb]
  \centering
  \begin{tabular}{c}
  \begin{lstlisting}[language=SQL]
    SELECT * FROM tbl WHERE tbl.str = "str"
  \end{lstlisting}
  \end{tabular}
  \caption[Example listing]{An example for a source code listing.}\label{fig:sample-listing}
\end{figure}
